Normalization by Evaluation for Sized Dependent Types

### 2017-02-24 nbe/icfp17.tex
[ ICFP 17 ] Correctness of algorithmic subtyping depending on whnormalization.

@@

@@

>>>>

\begin{corollary}[Type weak head normalization]
  If $\Gamma \der T$ then $T \evalsto W$ for some $W$.
\end{corollary}
\begin{proof}
  By the fundamental theorem of typing, $\RG \Gamma T {T\rhog}$,
  which implies $T \evalsto W$ by definition of the logical relation..
\end{proof}

@@

\section{Algorithmic subtyping}
\label{sec:asub}

@@

\begin{lemma}[Semantic subtyping implies algorithmic subtyping]
\label{lem:lrasub}
\bla
   If\/ $\LRTS \Gamma T A s$ and $\LRTS \Gamma {T'} {A'} {s'}$ and $A \leq A' \in \TYPE$ then $\Gamma \der T \asub T'$.
\end{lemma}
\begin{proof}
By induction on $A_1' \in s_1$ and $A_2' \in s_2$ and cases on the inductive relations
$\LRTS \Gamma {T_1'} {A_1'} {s_1}$ and $\LRTS \Gamma {T_2'} {A_2'} {s_2}$ and $A_1' \leq A_2' \in \TYPE$
we prove $\Gamma \der T_1' \asub T_2'$.

\nextcase $A_i' \evalsto N_i$ and $T_i' \evalsto n_i$ and $N_1 = N_2 \in \NE$ and $\RTY \Gamma {n_i} {N_i} {s_i}$.
By Cor.~\ref{cor:oneone} $\Gamma \der n_1 = n_2$, thus, by completeness of NbE, $\Nbe\Gamma n_1 \equiv \Nbe\Gamma n_2$, which entails $\Gamma \der T_1' \asub T_2'$.

\nextcase $A_i' \evalsto \piT {A_i} {B_i}$ and  $T_i' \evalsto \piT{U_i}{T_i}$.
From $\LRTS \Gamma {U_i} {A_i} {s_i}$ and $A_2 \leq A_1 \in \TYPE$ we get $\Gamma \der U_2 \asub U_1$
by induction hypothesis on $A_i \in s_i$.
Let $a := {\up{A_2} \var\Gamma}$.  With $\LR {\cext \Gamma {U_2}} {\ind 0} {(U_2\slift)} a {A_2}$ we get
$\LRTS {\cext \Gamma {U_2}} {T_i} {B_i[a]} s$ and $B_1[a] \leq B_2[a] \in \TYPE$.  Thus by induction hypothesis on $B_i[a] \in s_i$ we obtain $\cext \Gamma {U_2} \der T_1 \leq T_2$, together $\Gamma \der T_1' \asub T_2'$.

\begin{corollary}[Completeness of algorithmic subtyping]
\label{cor:complasub}
If $\Gamma \der T \leq T'$ then $\Gamma \der T \asub T'$.
\end{corollary}
\begin{proof}
  By the fundamental theorems $\RG \Gamma T {T\rhog}$ and $\RG \Gamma {T'} {T'\rhog}$ and $T\rhog \leq T'\rhog \in \TYPE$.
  By Lemma~\ref{lem:lrasub}, $\Gamma \der T \asub T'$.
\end{proof}

Since the algorithmic subtyping relation is equivalent to the declarative one, we can freely swap one relation for the other.

\begin{lemma}[Termination of algorithmic subtyping]
  \label{lem:termasub}
  If $\RG \Gamma T A$ and $\RG \Gamma {T'} {A'}$ then the query $\Gamma \der T \asub T'$ terminates.
\end{lemma}
\begin{proof}
  By induction on $A \in s$ and $A' \in s'$ and cases on $\RG \Gamma T A$ and $\RG \Gamma {T'} {A'}$.
\end{proof}

\begin{theorem}[Decidability of subtyping]
  If $\Gamma \der T,T'$, then $\Gamma \der T \leq T'$ is decided by the query $\Gamma \der T \asub T'$.
\end{theorem}
\begin{proof}
  By the fundamental theorem of typing, $\RG \Gamma T A$ and $\RG \Gamma {T'} {A'}$,
  thus, the query  $\Gamma \der T \asub T'$ terminates by Lemma~\ref{lem:termasub}.
  If successfully, then $\Gamma \der T \leq T'$ by soundness of algorithmic equality.
  If unsuccessfully, then $\Gamma \der T \leq T'$ is impossible by completeness of algorithmic equality.
\end{proof}
